Equalizer Programming Guide\\ {\vspace{-1ex}\footnotesize\sffamily\mdseries \htmladdnormallink{http://www.equalizergraphics.com/documents/Developer/ProgrammingGuide.pdf} {http://www.equalizergraphics.com/documents/ProgrammingGuide.pdf}}

### 2007-10-28 Developer/ProgrammingGuide/paper.tex
Update Programming Guide and website for 0.4 release
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 0.9.4 for Equalizer 0.4\\[\medskipamount]
  \today
}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    0.9.4   & Oct 26, 2007 & release candidate 1\\
    0.9.3   & Oct 21, 2007 & update title page, add front matter\\
    0.9.2   & Oct 19, 2007 & Jonas' feedback, changes related to
                             per-node frame sync\\
    0.9.1   & Oct 8,  2007 & review pass\\
    0.9     & Oct 7,  2007 & add eqHello, draft assembly in eVolve\\
    0.8     & Sep 28, 2007 & add head tracking, finish channel,
                             proof-reading pass\\
    0.7     & Sep 21, 2007 & start channel and image compositing,
                             add event handling, move to new eqPly
                             code base\\
    0.6     & Sep 14, 2007 & add pipe, window and object manager\\
    0.5     & Sep 3,  2007 & add config and partly pipe\\
    0.4     & Aug 31, 2007 & add distributed objects\\
    0.3     & Aug 26, 2007 & add application and render client\\
    0.2     & Aug 20, 2007 & add main function\\
    0.1     & Aug 19, 2007 & outline the basic concepts\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

Equalizer provides a framework for the development of parallel OpenGL
applications. Equalizer-based applications can run from a single
shared-memory system with a one or multiple graphics cards up to
large-scale graphics clusters. This Programming Guide introduces the
programming interface, often using the Equalizer \textsf{eqPly} example
as a guideline.

@@

The main thread is responsible for maintaining the application logic. It
reacts on user events, updates the data model and requests new frames to
be rendered. It drives the whole application, as shown in \fig{model}

@@

Rendering threads on a single node are by default synchronized. When a
frame is finished, all local rendering threads are done
drawing. Therefore the application can safely modify the data between
the end of a frame and the beginning of a new frame. Furthermore, only
one instance of the application data has to be maintained within a
process since all rendering threads are quaranteed to draw the same frame.

@@

@@

\fig{fConfig} shows one example configuration for a four-side
CAVE, running on two machines (nodes) using three graphics
cards (pipes) with one window each to render to the four output channels
connected to the projectors for each of the walls. The compound
description is only used by the server to compute the rendering
tasks. The application is not aware of compounds, and does not need to
concern itself with the parallel rendering logics of a configuration.

@@

An \textsf{eq::Node} is the representation of a single computer in a
cluster. One operating system process of the render client will be used
for each node. Each configuration might also use an application node, in
which case the application process is also used for rendering. All
node-specific task methods are executed from the main application
thread.

@@

The \textsf{eq::Pipe} is the abstraction of a graphics card (GPU).  In
the current implementation it is also one operating system
thread. Non-threaded pipes are supported for integrating with
thread-unsafe libraries, but have various performance caveats. All pipe,
window and channel task methods are executed from the pipe's thread, or
in the case of non-threaded pipes from the main application
thread\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/nonthreaded.html}}.

@@

An \textsf{eq::Window} holds a drawable and an OpenGL context. The
drawable can be an on-screen window or an off-screen\footnote{Off-screen drawables are not implemented yet, but can be
  created by the application and used with Equalizer} PBuffer or
framebuffer object (FBO). The window holds
window-system-specific handles to the drawable and context, e.g. an X11
window \textsf{XID} and \textsf{GLXContext} for the glX window system.

@@

The \textsf{eq::Channel} is the abstraction of an OpenGL viewport within
its parent window. It is the entity executing the actual rendering. The
channel's viewport might be overwritten when it is rendering for another
channel during scalable rendering.

@@

\subsubsection{Compound Channels}
Each compound has a channel, which it uses to execute the rendering
tasks. One channel might be used by multiple compounds. Unused channels
are not instantiated during initialization. The rendering tasks for the
channels are computed by the server and send to the appropriate render
clients.

@@

\subsubsection{Decomposition - Attributes}
Compounds have attributes which configure the decomposition of the
destination's channel viewport, frustum and database. A
\textsf{viewport} decomposes the destination channel and frustum in
screen space. A \textsf{range} tells the application to render a part of
its database, and an \textsf{eye} pass can selectively render different
stereo passes. Setting one or multiple attributes causes the parent's
view to be decomposed accordingly. Attributes are cumulative, that is,
intermediate compound attributes affect and therefore decompose the
rendering of all their children.

@@

In this section the source code of \textsf{eqPly} is discussed in
detail, and relevant design decision, caveats and other remarks are
raised.

@@

The \textsf{eqPly} classes fall into two categories: Subclasses of the
rendering entities introduced in \sref{sResource}, and two other classes
for distributing data. The function and typical usage for each of the
rendering entities is discussed in this section.

@@

The local initialization of a node creates a local listening socket, so
that the node. Therefore the \textsf{eq::Client} can communicate over
the network with other nodes, such as the server and the rendering
clients.

@@

\begin{enumerate}
\item \textbf{Application}: The executable started by the user, the
  controlling entity in the rendering session.
\item \textbf{Auto-launched render client}: The typical render client,
  started by the server. The server starts the executable with special
  parameters, which cause \textsf{Client::initLocal} to never
  return. During exit, the server terminates the process. By default,
  the server starts render client using \textsf{ssh}.
\item \textbf{Resident render client}: Manually pre-started render
  client, listening on a specified port for server commands. This mode
  is selected using the command-line option \textsf{--eq-client} and
  potentially \textsf{--eq-listen \textless address\textgreater} and
  \textsf{-r}\footnote{see
    \link{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
\end{enumerate}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \subfigure[]{\label{fSync}
    \includegraphics[width=.25\textwidth]{images/sync}}\hfil
  \subfigure[]{\label{fAsync}
    \includegraphics[width=.25\textwidth]{images/async}}%
  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{wrapfigure}
\fig{fSyncAsync} shows the execution of (hypothetical) rendering tasks
without latency (\fig{fSync}) and with a latency of
one (\fig{fAsync}). With \textsf{eqPly}, a speedup of 15\% has been
observed on a five-node rendering cluster when using a latency of one
instead of no
latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
latency of one or two frames is normally not perceived by the user.

When the main loop is finished, \textsf{Config::finishAllFrames} catches
up with the latency. It returns after all outstanding frames have been
rendered, and is needed to provide an accurate measurement of the
framerate:

@@

The \textsf{InitData} class contains a string of variable
length. Therefore it uses the second approach of manually serializing
and deserializing its data. Serialization in \textsf{getInstanceData}
and deserialization in \textsf{applyInstanceData} is performed by
streaming all member variable to or from the provided data
streams. Efficient buffering and data transport between nodes is
implemented in the data streams:

@@

The data input and output streams perform no type checking on the data.
It is the application's responsibility to exactly match the order and
types of variables during serialization and deserialization.

@@

In addition to the instance data (de)serialization methods used to map
an object, versioned objects may implement \textsf{pack} and
\textsf{unpack} to serialize or deserialize the changes since the last
version. These methods can

@@

The config initialization happens in parallel, that is, all config
initialization tasks are transmitted by the server at once and their
execution is synchronized later.

@@

The frame data is registered before the initialization data, since its
identifier is transmitted using the \textsf{InitData}. The identifier of
the initialization data is transmitted to the render client nodes using
the \textsf{initID} parameter of \textsf{eq::Config::init}. Equalizer
will pass this identifier to all \textsf{configInit} calls of the
respective objects:

@@

The exit function of the configuration stops the render clients by calling
\textsf{eq::Con\-fig::exit}, and then deregisters the initialization and
frame data objects with the session:

@@

If a tracker is used, the current head position and orientation is
retrieved and given to Equalizer, which uses the head matrix together
with the wall or projection description to compute the view
frustra\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/immersive.html}}.

@@

\fig{fFrameSync} outlines the synchronization for the application, node
and pipe classes. Please note that \textsf{Config::finishFrame} does
block until the current frame has been released locally and until the
frame \textsf{current - latency} has been released by all nodes. The
window and channel synchronization is similar and omitted for
simplicity.

@@

Pipe threads are not explicitely synchronized with each other, that is,
pipes might be rendering different frames at one given time. Therefore
frame-specific data has to be allocated for each pipe thread, which in
the \textsf{eqPly} example is the frame data. The frame data is a member
variable of the \textsf{eqPly::Pipe}, and is mapped to the identifier
provided by the initialization data. The initialization in
\textsf{eq::Pipe} does GPU-specific initialization, which is
window-system-dependent. On AGL the display ID is determined, and on glX
the display connection, to be used by the child windows, is opened.

@@

Each pipe might use a different window system for rendering, which is
determined before \textsf{Pipe::configInit} by
\textsf{Pipe::selectWindowSystem}. The default implementation of
\textsf{selectWindowSystem} loops over all window systems and returns
the first supported window system, determined using
\textsf{supportsWindowSystem}.

The \textsf{eqPly} examples allows selecting the window system using a
command line option. Therefore the implementation of
\textsf{selectWindowSystem} is overwritten, and returns the specified
window system, if it is supported:

@@

The Equalizer window holds an OpenGL drawable and rendering
context. When using the default window initialization functions, all
windows of a pipe share the OpenGL context. This allows reuse of OpenGL
objects such as display lists and textures between all windows of one
pipe.

@@

The initialization sequence uses multiple, overrideable task
methods. The main task method \textsf{configInit} executes a `child'
task method to create the drawable and context. The child task method
depends on the window system of the pipe. The default implementations of
\textsf{configInitGLX}, \textsf{configInitWGL} or \textsf{configInitAGL}
create an on-screen window using OS-specific methods. If the OpenGL
drawable and context was created successfully, \textsf{configInit} calls
\textsf{configInitGL}, which will perform the generic OpenGL
state initialization. The default implementation sets up some typical
OpenGL state, e.g., it enables the depth test.

@@

The window exit function deallocates all OpenGL objects when the state object is
about to be disposed. The object manager does not delete the object in
its destructor, since it does not know if an OpenGL context is still
current. Additionally, \textsf{eq::Window::configExit} is called to
destroy the drawable and context:

@@

The object manager is not strictly a part of the window. It
is discussed here since the \textsf{eqPly} window uses an object manager.

@@

The OM is a template class. The template type is the key which is used to
identify objects. The same key is used by all contexts to get the
OpenGL name of an object. In \textsf{eqPly}, a key of type \textsf{const
  void *} is used. The rendering code uses the address of the data item
to be rendered as the key to obtain the associated OpenGL object.

@@

This section discusses some additional important features not covered
by the previous \textsf{eqPly} section. Where possible, code examples
from the Equalizer distribution are used.

@@

Event handling requires a lot of flexibility. On one hand, the
implementation differs slightly for each operating and window system due
to conceptual differences in the OS-specific implementation. On the
other hand, each application and widget set has its own model on how
events are to be handled. Therefore, event handling in Equalizer is
customizable in any stage of the processing, to the extreme of making it
possible to disable all event-related code in Equalizer. In this aspect,
Equalizer substantially differs from GLUT, which imposes an event model
and hides most of the event handling in \textsf{glutMainLoop}. More
information on event handling can be found on the Equalizer
website\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/eventHandling.html}}.

@@

Where possible, events are received and processed by a separate per-node
event thread to allow asynchronous\footnote{with respect to the
  rendering} event handling. Currently an event thread is only used by
the X11/glX window system. WGL receives and processes the events from
the pipe threads that created the windows. AGL receives the events from
application or node main thread. Whenever the term \textbf{event thread}
is used, it refers to the thread receiving the event, such as a per-node
thread for glX, the pipe thread for WGL and the main thread for AGL.

@@

Event handling is initialized whenever a new, window-system-specific
pipe or window handle is set. First, \textsf{exitEventHandler} is called
to de-initialize event handling for the old handle (if set), and then
\textsf{initEvent\-Handler} is called for the new handle.

@@

The \textsf{Config::sendEvent} method transmits a
\textsf{eq::ConfigEvent} or derived class to the application. The
ConfigEvent has to be a C-type structure, and its \textsf{size}
member has to be set to the full size of the event to be transmitted.
Each event has a type which is used to identify it by the config 
processing function.

@@

On the application end, \textsf{Config::handleEvent} uses the ostream
operator for the derived config event to output these events in a nicely
formatted way:

@@

Normally the programmer does not need to interfere with the image
compositing, or only at a high level, for example to order the input
frames or to optimize the readback. The following sections provide a
detailed description of the image compositing API in Equalizer.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
\end{wrapfigure}
The main idea for direct send is to parallelize the costly recomposition
for database (sort-last) decomposition. With each additional source
channel, the amount of pixel data to be composited grows linearly. When
using the naive approach of compositing all frames on the destination
channel, this channel quickly becomes the bottleneck in the
system. Direct send distributes this workload evenly across all source
channels, and thereby keeps the compositing work per channel constant.

@@

\fig{fDirectSend} illustrates this algorithm for three channels. The
Equalizer website contains a presentation for this
algorithm\footnote{\link{http://www.equalizergraphics.com/documents/EGPGV07.pdf}}.

@@

For output frames, the frame data's pixel viewport (pvp) is the area of
the frame buffer to read back. It will transport the offset from the
source to the destination channel, that is, the frame data pvp for input
frames position the pixel data on the destination. This has the effect
that a partial framebuffer readback will end up in the same place in the
destination channels.

@@

Conceptually, the individual volume bricks of each of the source
channels produces pixel data which can be handled like one big 'slice'
through the full texture. Therefore they have to be blended
back-to-front in the same way as the slice planes during rendering.

DB compounds have the advantage of scaling any part of the volume
rendering pipeline: texture and main memory (smaller bricks for each
channel), fill rate (less samples per channel) and IO bandwidth for
time-dependend data (less data per time step and channel). Since the
amount of texture memory needed for each node decreases linearly, they
make it possible to render data sets which are not feasible to
visualize with any other approach.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \vspace{-4ex}\includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
In order to compute the assembly order, \textsf{eVolve} uses the angle
between the origin$\rightarrow$slice vector and the near plane, as shown
in \fig{fBlendPerspective}. When the angle becomes bigger than
90\textdegree, the compositing order of the remaining frames has to be
changed. The result image of this composition naturally looks the same
as the volume rendering would when rendered on a single
channel. \fig{fVolResult} shows the result of the composition from
\fig{fBlendPerspective}.

@@

\fig{fMono} illustrates a monoscopic view frustum. The viewer is
positioned at the origin, and the frustum is completely symmetric. This
is the typical view frustum for non-stereoscopic applications.

>>>>

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.0 for Equalizer 0.4\\[\medskipamount]
  \today
}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

Equalizer provides a framework for the development of parallel OpenGL
applications. Equalizer-based applications can run from a single
shared-memory system with one or multiple graphics cards up to
large-scale graphics clusters. This Programming Guide introduces the
programming interface, often using the Equalizer \textsf{eqPly} example
as a guideline.

@@

The main thread is responsible for maintaining the application logic. It
reacts on user events, updates the data model and requests new frames to
be rendered. It drives the whole application, as shown in \fig{fModel}.

@@

Rendering threads on a single node are by default synchronized. When a
frame is finished, all local rendering threads are done
drawing. Therefore the application can safely modify the data between
the end of a frame and the beginning of a new frame. Furthermore, only
one instance of the application data has to be maintained within a
process since all rendering threads are guaranteed to draw the same frame.

@@

\subsection{\label{sConfig}Config}

@@

\fig{fConfig} shows an example configuration for a four-side
CAVE, running on two machines (nodes) using three graphics
cards (pipes) with one window each to render to the four output channels
connected to the projectors for each of the walls. The compound
description is only used by the server to compute the rendering
tasks. The application is not aware of compounds, and does not need to
concern itself with the parallel rendering logics of a configuration.

@@

The \textsf{eq::Node} class is the representation of a single computer
in a cluster. One operating system process of the render client will be
used for each node. Each configuration might also use an application
node, in which case the application process is also used for
rendering. All node-specific task methods are executed from the main
application thread.

@@

The \textsf{eq::Pipe} class is the abstraction of a graphics card (GPU).
In the current implementation it is also one operating system
thread. Non-threaded pipes are supported for integrating with
thread-unsafe libraries, but have various performance caveats. All pipe,
window and channel task methods are executed from the pipe's thread, or
in the case of non-threaded pipes from the main application
thread\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/nonthreaded.html}}.

@@

The \textsf{eq::Window} class holds a drawable and an OpenGL
context. The drawable can be an on-screen window or an
off-screen\footnote{Off-screen drawables are not implemented yet, but
  can be created by the application and used with Equalizer} PBuffer or
framebuffer object (FBO). The window holds window-system-specific
handles to the drawable and context, e.g., an X11 window \textsf{XID} and
\textsf{GLXContext} for the glX window system.

@@

The \textsf{eq::Channel} class is the abstraction of an OpenGL viewport
within its parent window. It is the entity executing the actual
rendering. The channel's viewport might be overwritten when it is
rendering for another channel during scalable rendering.

@@

\subsubsection{Compound Channels}
Each compound has a channel, which is used by the compound to execute
the rendering tasks. One channel might be used by multiple
compounds. Unused channels are not instantiated during
initialization. The rendering tasks for the channels are computed by the
server and send to the appropriate render clients.

@@

\subsubsection{Decomposition - Attributes}
Compounds have attributes which configure the decomposition of the
destination's channel viewport, frustum and database. A
\textsf{viewport} decomposes the destination channel and frustum in
screen space. A \textsf{range} tells the application to render a part of
its database, and an \textsf{eye} rendering pass can selectively render
different stereo passes. Setting one or multiple attributes causes the
parent's view to be decomposed accordingly. Attributes are cumulative,
that is, intermediate compound attributes affect and therefore decompose
the rendering of all their children.

@@

In this section the source code of \textsf{eqPly} is explained in
detail, and relevant design decisions, caveats and other remarks are
raised.

@@

The \textsf{eqPly} classes fall into two categories: Subclasses of the
rendering entities introduced in \sref{sConfig}, and two other classes
for distributing data. The function and typical usage for each of the
rendering entities is discussed in this section.

@@

The local initialization of a node creates a local listening socket,
which allows the \textsf{eq::Client} to communicate over the network
with other nodes, such as the server and the rendering clients.

@@

\begin{enumerate}
\item \textbf{Application}: The executable started by the user, the
  controlling entity in the rendering session.
\item \textbf{Auto-launched render client}: The typical render client,
  started by the server. The server starts the executable with special
  parameters, which cause \textsf{Client::initLocal} to never
  return. During exit, the server terminates the process. By default,
  the server starts the render client using \textsf{ssh}.
\item \textbf{Resident render client}: Manually pre-started render
  client, listening on a specified port for server commands. This mode
  is selected using the command-line option \textsf{--eq-client} and
  potentially \textsf{--eq-listen \textless address\textgreater} and
  \textsf{-r}\footnote{see
    \link{http://www.equalizergraphics.com/documents/design/residentNodes.html}}.
\end{enumerate}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \subfigure[]{\label{fSync}
    \includegraphics[width=.25\textwidth]{images/sync}}\hfil
  \subfigure[]{\label{fAsync}
    \includegraphics[width=.25\textwidth]{images/async}}%
  {\caption{\small\label{fSyncAsync}Synchronous and asynchronous execution}}
\end{wrapfigure}
\fig{fSyncAsync} shows the execution of (hypothetical) rendering tasks
without latency (\fig{fSync}) and with a latency of
one frame (\fig{fAsync}). With \textsf{eqPly}, a speedup of 15\% has been
observed on a five-node rendering cluster when using a latency of one frame
instead of no
latency\footnote{\link{http://www.equalizergraphics.com/scalability.html}}. A
latency of one or two frames is normally not perceived by the user.

When the main rendering loop has finished,
\textsf{Config::finishAllFrames} is called to catch up with the
latency. It returns after all outstanding frames have been rendered, and
is needed to provide an accurate measurement of the framerate:

@@

The \textsf{InitData} class contains a string of variable
length. Therefore it uses the second approach of manually serializing
and de-serializing its data. Serialization in \textsf{getInstanceData}
and de-serialization in \textsf{applyInstanceData} is performed by
streaming all member variable to or from the provided data
streams. Efficient buffering and data transport between nodes is
implemented in the data streams:

@@

The data input and output streams perform no type checking on the data.
It is the application's responsibility to exactly match the order and
types of variables during serialization and de-serialization.

@@

In addition to the instance data (de)serialization methods used to map
an object, versioned objects may implement \textsf{pack} and
\textsf{unpack} to serialize or de-serialize the changes since the last
version.

@@

The config initialization happens in parallel, that is, all config
initialization tasks are transmitted by the server at once and their
completion is synchronized afterwards.

@@

The frame data is registered before the initialization data, since its
identifier is transmitted using the \textsf{InitData}. The identifier of
the initialization data is transmitted to the render client nodes using
the \textsf{initID} parameter of \textsf{eq::Config::init}.

Equalizer will pass this identifier to all \textsf{configInit} calls of
the respective objects:

@@

The exit function of the configuration stops the render clients by calling
\textsf{eq::Con\-fig::exit}, and then de-registers the initialization and
frame data objects with the session:

@@

If a tracker is used, the current head position and orientation is
retrieved and passed to Equalizer, which uses the head matrix together
with the wall or projection description to compute the view
frustra\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/immersive.html}}.

@@

\fig{fFrameSync} outlines the synchronization for the application, node
and pipe classes. Please note that \textsf{Config::finishFrame} does
block until the current frame has been released locally and until the
frame \textsf{current - latency} has been released by all nodes. The
window and channel synchronization are similar and omitted for
simplicity.

@@

Pipe threads are not explicitely synchronized with each other, that is,
pipes might be rendering different frames at any given time. Therefore
frame-specific data has to be allocated for each pipe thread, which in
the \textsf{eqPly} example is the frame data. The frame data is a member
variable of the \textsf{eqPly::Pipe}, and is mapped to the identifier
provided by the initialization data. The initialization in
\textsf{eq::Pipe} does the GPU-specific initialization, which is
window-system-dependent. On AGL the display ID is determined, and on glX
the display connection is opened.

@@

Each pipe might use a different window system for rendering, which is
determined before \textsf{Pipe::configInit} by
\textsf{Pipe::selectWindowSystem}. The default implementation of
\textsf{selectWindowSystem} loops over all window systems and returns
the first supported window system, determined by using
\textsf{supportsWindowSystem}.

The \textsf{eqPly} examples allows selecting the window system using a
command line option. Therefore the implementation of
\textsf{selectWindowSystem} is overwritten and returns the specified
window system, if supported:

@@

The Equalizer window holds an OpenGL drawable and a rendering
context. When using the default window initialization functions, all
windows of a pipe share the OpenGL context. This allows reuse of OpenGL
objects such as display lists and textures between all windows of one
pipe.

@@

The initialization sequence uses multiple, overrideable task
methods. The main task method \textsf{configInit} executes a `child'
task method to create the drawable and context. The child task method
depends on the window system of the pipe. The default implementations of
\textsf{configInitGLX}, \textsf{configInitWGL} or \textsf{configInitAGL}
create an on-screen window using OS-specific methods. If the OpenGL
drawable and context were created successfully, \textsf{configInit} calls
\textsf{configInitGL}, which performs the generic OpenGL
state initialization. The default implementation sets up some typical
OpenGL state, e.g., it enables the depth test.

@@

The window exit function de-allocates all OpenGL objects when the state
object is about to be disposed. The object manager does not delete the
object in its destructor, since it does not know if an OpenGL context is
still current. Additionally, \textsf{eq::Window::configExit} is called
to destroy the drawable and context:

@@

The object manager is not strictly a part of the window. It
is mentioned here since the \textsf{eqPly} window uses an object manager.

@@

The OM is a template class. The template type is the key used to
identify objects. The same key is used by all contexts to get the
OpenGL name of an object. In \textsf{eqPly}, a key of type \textsf{const
  void *} is used. The rendering code uses the address of the data item
to be rendered as the key to obtain the associated OpenGL object.

@@

This section discusses some additional important features not covered
by the previous \textsf{eqPly} section. Where possible code examples
from the Equalizer distribution are used.

@@

Event handling requires a lot of flexibility. On one hand, the
implementation differs slightly for each operating and window system due
to conceptual differences in the OS-specific implementation. On the
other hand, each application and widget set has its own model on how
events are to be handled. Therefore, event handling in Equalizer is
customizable at any stage of the processing, to the extreme of making it
possible to disable all event-related code in Equalizer. In this aspect,
Equalizer substantially differs from GLUT, which imposes an event model
and hides most of the event handling in \textsf{glutMainLoop}. More
information on event handling can be found on the Equalizer
website\footnote{see
  \link{http://www.equalizergraphics.com/documents/design/eventHandling.html}}.

@@

Where possible events are received and processed by a separate per-node
event thread to allow asynchronous\footnote{with respect to the
  rendering} event handling. Currently an event thread is only used by
the X11/glX window system. WGL receives and processes the events from
the pipe threads that created the windows. AGL receives the events from
application or node main thread. Whenever the term \textbf{event thread}
is used, it refers to the thread receiving the event, such as a per-node
thread for glX, the pipe thread for WGL and the main thread for AGL.

@@

Event handling is initialized whenever a new window-system-specific
pipe or window handle is set. First, \textsf{exitEventHandler} is called
to de-initialize event handling for the old handle (if set), and then
\textsf{initEvent\-Handler} is called for the new handle.

@@

The \textsf{Config::sendEvent} method transmits an
\textsf{eq::ConfigEvent} or derived class to the application. The
ConfigEvent has to be a C-type structure, and its \textsf{size}
member has to be set to the full size of the event to be transmitted.
Each event has a type which is used to identify it by the config 
processing function.

@@

On the application end, \textsf{Config::handleEvent} uses the
\textsf{ostream} operator for the derived config event to output these
events in a nicely formatted way:

@@

Normally the programmer does not need to interfere with the image
compositing. Changes are sometimes required at a high level, for example
to order the input frames or to optimize the readback. The following
sections provide a detailed description of the image compositing API in
Equalizer.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
\end{wrapfigure}
The main idea for direct send is to parallelize the costly recomposition
for database (sort-last) decomposition. With each additional source
channel, the amount of pixel data to be composited grows linearly. When
using the simple approach of compositing all frames on the destination
channel, this channel quickly becomes the bottleneck in the
system. Direct send distributes this workload evenly across all source
channels, and thereby keeps the compositing work per channel constant.

@@

\fig{fDirectSend} illustrates this algorithm for three channels. The
Equalizer website contains a presentation explaining this
algorithm\footnote{\link{http://www.equalizergraphics.com/documents/EGPGV07.pdf}}.

@@

For output frames, the frame data's pixel viewport is the area of the
frame buffer to read back. It will transport the offset from the source
to the destination channel, that is, the frame data pixel viewport for
input frames position the pixel data on the destination. This has the
effect that a partial framebuffer readback will end up in the same place
in the destination channels.

@@

Conceptually, the individual volume bricks of each of the source
channels produces pixel data which can be handled like one big 'slice'
through the full texture. Therefore they have to be blen\-ded
back-to-front in the same way as the slice planes during rendering.

DB compounds have the advantage of scaling any part of the volume
rendering pipeline: texture and main memory (smaller bricks for each
channel), fill rate (less samples per channel) and IO bandwidth for
time-dependent data (less data per time step and channel). Since the
amount of texture memory needed for each node decreases linearly, they
make it possible to render data sets which are not feasible to
visualize with any other approach.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \vspace{-4ex}\includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
In order to compute the assembly order, \textsf{eVolve} uses the angle
between the origin$\rightarrow$slice vector and the near plane, as shown
in \fig{fBlendPerspective}. When the angle becomes greater than
90\textdegree, the compositing order of the remaining frames has to be
changed. The result image of this composition naturally looks the same
as the volume rendering would when rendered on a single
channel. \fig{fVolResult} shows the result of the composition from
\fig{fBlendPerspective}.

@@

\fig{fMono} illustrates a monoscopic view frustum. The viewer is
positioned at the origin of the global coordinate system, and the
frustum is completely symmetric. This is the typical view frustum for
non-stereoscopic applications.
